---
title: 'BioinfoSummer2015: 450k Analysis Workshop'
author: "Jovana Maksimovic and Belinda Phipson"
date: "27 November 2015"
output: 
  html_document: 
    fig_caption: yes
    fig_height: 7
    fig_width: 10
    number_sections: yes
    toc: yes
---

#Introduction
Methylation, the addition of a methyl group to a CG dinucleotide of the DNA, is the most extensively studied epigenetic mark. Although it can be measured in many ways, the most popular platform for measuring methylation has been the Illumina HumanMethylation450 (450k) array. In this workshop, we will introduce the features of the 450k array and how to analyse the methylation array data using R. We will cover the steps involved in the standard 450k analysis pipeline including: quality control, filtering, normalization, data exploration and probe-wise differential methylation array analysis. We will also cover other approaches such as differential methylation analysis of regions, differential variability analysis and gene ontology analysis. Finally, we will provide some examples of how to visualise methylation array data.

##Preparing for the workshop
###Software
To participate in this workshop you will need the following software installed on your computer:

- [R (3.2.2)](https://cran.r-project.org/)

- [RStudio (0.99.489)](https://www.rstudio.com/products/rstudio/download/)

You should then start **RStudio** and install [Bioconductor (3.2)](http://bioconductor.org/install/) and several required packages by running the following commands:

```{r, eval=FALSE}
## install required Bioconductor packages
source("http://www.bioconductor.org/biocLite.R")
biocLite()
biocLite(c("minfi","limma","IlluminaHumanMethylation450kanno.ilmn12.hg19","minfiData",
          "IlluminaHumanMethylation450kmanifest","RColorBrewer","missMethyl",
          "DMRcate","stringi","plyr","Gviz","matrixStats","GO.db","BiasedUrn"))
```

###Data
To demonstrate the various aspects of analysing methylation data, we will be using a small methylation dataset that has been previously analysed and published. The data is of the methylation profiles of various sorted T-cell types. More information on the samples and the dataset can be found in the original [paper](http://www.bloodjournal.org/content/122/16/2823?sso-checked=true). The data can be downloaded from the [MCRI Owncloud](https://owncloud.mcri.edu.au/index.php/s/zESg9ejgf065VOy) (password: *biosummer2015*) until Dec 12th. Please download all of the data files into the **same** directory.

## Illumina HumanMethylation450 BeadChips
[Illumina HumanMethylation450 (450k) arrays](http://www.illumina.com/products/methylation_450_beadchip_kits.html) are a cost-effective alternative to whole genome bisulfite sequencing, and as such have been widely used to profile DNA methylation, particularly for studies with large numbers of samples. Illumina have recently release a new platform, the [Illumina MethylationEPIC array](http://www.illumina.com/products/infinium-methylation-epic-array.html), which increases the number of CpG probes from ~450,000 to ~850,000. Although the number of CpGs covered is increased on the new array, the same probe chemistry is being used and thus the principles for analysing the 450k array should be largely applicable to the new 850k array.     

# Standard analysis
This workshop will cover various aspects of a basic differential methylation analysis of 450k arrays. There are many R Bioconductor packages available for analysing 450k array methylation data, including [minfi](http://bioconductor.org/packages/release/bioc/html/minfi.html), [missMethyl](http://www.bioconductor.org/packages/release/bioc/html/missMethyl.html), [wateRmelon](https://www.bioconductor.org/packages/release/bioc/html/wateRmelon.html) and [methylumi](https://www.bioconductor.org/packages/release/bioc/html/methylumi.html). We will begin with an example of a **probe-wise** methylation analysis using *minfi* and *limma*. By **probe-wise** analysis we mean that we will obtain a moderated t-statistic and p-value for each individual CpG probe, which we can use to determine which individual CpGs are differentially methylated at some significance level for our comparisons of interest.

### Loading the data
We will begin by loading all the package libraries that will be required for the analysis.
```{r, cache=TRUE}
## load packages required for analysis
library(limma)
library(minfi)
library(IlluminaHumanMethylation450kanno.ilmn12.hg19)
library(IlluminaHumanMethylation450kmanifest)
library(RColorBrewer)
library(missMethyl)
library(matrixStats)
library(minfiData)
library(Gviz)
library(DMRcate)
```

Load the Illumina manifest which contains all of the annotation information for each of the CpG probes on the 450k array. This will help us determine where our differentially methylated probes are located in a genomic context. 
```{r, cache=TRUE}
## get the 450k annotation data
ann450k = getAnnotation(IlluminaHumanMethylation450kanno.ilmn12.hg19)
head(ann450k)
```

Next, we will load in the data. The sample sheet file is essentially a CSV (comma-separated) file containing one line per sample, with a number of columns describing each sample. The format expected by the `read.450k.sheet` function is based on the sample sheet file Illumina provides. It is also very similar to the targets file made popular by the *limma* package (see the extensive [Limma User's Guide](https://www.bioconductor.org/packages/release/bioc/vignettes/limma/inst/doc/usersguide.pdf)). Raeding in the sample sheet essentially creates a `data.frame`. If they exist, a column named `Sentrix_Position` is renamed to `Array` and `Sentrix_ID` is renamed to `Slide`. The function also attempts to derive the path to the IDAT files using the `baseDir` and the `Slide` and `Array` information. This will be stored in a column named `Basename`.
```{r, cache=TRUE}
## set up a path to the directory you donloaded the data to
dataDirectory = "D:/Work/BioinfoSummer2015/450kAnalysisWorkshop/data"

## read in the sample sheet for the experiment
targets = read.450k.sheet(dataDirectory, pattern="SampleSheet.csv")
targets
```

Now that we have imported the information about what and where all the data is, we can load the raw methylation data into R from the IDAT files. We can then rename the samples with more descriptive names and calculate detection p-values, which will allow us to identify poor quality probes.
```{r, cache=TRUE}
## read in the raw data from the IDAT files
rgSet = read.450k.exp(targets=targets)
rgSet

## give the samples meaningful names
targets$ID = paste(targets$Sample_Group,targets$Sample_Name,sep=".")
sampleNames(rgSet) = targets$ID
rgSet

## calculate the detection P-values
detP = detectionP(rgSet)
head(detP)
```

### Quality control
Plotting the mean detection p-value for each sample will allow us to gauge whether any samples have many failed probes - this will be indicated by a large mean detection p-value. Samples with mean detection p-values exceeding a cutoff such as o.05 can be excluded from further analysis. 
```{r, cache=TRUE}
pal = brewer.pal(8,"Dark2")
## look at the mean detection P-values across all samples to identify any failed samples
par(mfrow=c(1,2))
barplot(colMeans(detP),col=pal[factor(targets$Sample_Group)],las=2,cex.names=0.8,
        main="Mean detection p-values")
abline(h=0.05,col="red")
legend("topleft",legend=levels(factor(targets$Sample_Group)),fill=pal,bg="white")

barplot(colMeans(detP),col=pal[factor(targets$Sample_Group)],las=2,cex.names=0.8,ylim=c(0,0.002),
        main="Mean detection p-values")
abline(h=0.05,col="red")
legend("topleft",legend=levels(factor(targets$Sample_Group)),fill=pal,bg="white")

## remove failed samples
keep = colMeans(detP) < 0.05
rgSet = rgSet[,keep]
targets = targets[keep,]
detP = detP[,keep]

rgSet

targets
```

You can also generate other quality control plots using the *minfi* `qcReport` function. The *minfi* [vignette](http://bioconductor.org/packages/release/bioc/vignettes/minfi/inst/doc/minfi.pdf) describes the various plots in this report and how they should be interpreted. Generally, samples that look poor based on mean detection p-value will also look poor using other metrics and it is usually advisable to exclude them from further analysis.
```{r, eval=FALSE}
## produce a general QC report for the data
qcReport(rgSet,sampNames=targets$ID,sampGroups=targets$Sample_Group,pdf="qcReport.pdf")
```

### Normalisation
To minimise the unwanted variation within and between samples, various data normalizations can be applied. As 450k arrays use 2 different probe types, Infinium I and II, on the same array, studies suggest that it is advisable to perform within-array normalization between the 2 probe types. Normalization can also be performed between arrays. Here we provide examples of 2 different normalization types: SWAN (within-array only) and SQN (within and between array). 
```{r, cache=TRUE}
## create a MethylSet object from the raw data
mSetRaw = preprocessRaw(rgSet)
mSetRaw

## normalize the data
mSetSw = preprocessSWAN(rgSet = rgSet, mSet = mSetRaw, verbose=TRUE) ## SWAN method
mSetSw = mapToGenome(mSetSw)

mSetSq = preprocessQuantile(rgSet) ## SQN method

par(mfrow=c(1,3))
densityPlot(rgSet, sampGroups = targets$Sample_Group,main="Raw")
densityPlot(getBeta(mSetSw), sampGroups = targets$Sample_Group,main="SWAN")
densityPlot(getBeta(mSetSq), sampGroups = targets$Sample_Group,main="SQN")
```

### Data exploration
Multi dimensional scaling (MDS) plots are excellent for visualising your data, and are usually one of the first kinds of plots you should make when exploring your data. It is based on principle componenents analysis and is an unsupervised method for looking at the similarities and differences between the samples of the data. Samples that are more similar to each other should cluster together, and samples that are very different should be further apart on the plot. Dimension 1 (or principle component 1) captures the greatest source of variation in the data, dimension 2 captures the second greatest source of variation in the data and so on.  Colouring in the data points or labels by known factors of interest can often pull out exactly what the greatest sources of variation are in the data. It's also possible to use MDSplots for figuring out sample mix-ups.
```{r, cache=TRUE}
## MDS plots to look at largets sources of variation
par(mfrow=c(1,2))
plotMDS(getM(mSetSq), top=1000, gene.selection = "common", col=pal[factor(targets$Sample_Group)])
legend("topright",legend=levels(factor(targets$Sample_Group)),text.col=pal)

plotMDS(getM(mSetSq), top=1000, gene.selection = "common", col=pal[factor(targets$Sample_Source)])
legend("topright",legend=levels(factor(targets$Sample_Source)),text.col=pal)

par(mfrow=c(1,1))
## Examine higher dimensions to look at other sources of variation
plotMDS(getM(mSetSq), top=1000, gene.selection = "common", col=pal[factor(targets$Sample_Group)], dim=c(1,3))
legend("topright",legend=levels(factor(targets$Sample_Group)),text.col=pal)

plotMDS(getM(mSetSq), top=1000, gene.selection = "common", col=pal[factor(targets$Sample_Group)], dim=c(2,3))
legend("topright",legend=levels(factor(targets$Sample_Group)),text.col=pal)

plotMDS(getM(mSetSq), top=1000, gene.selection = "common", col=pal[factor(targets$Sample_Group)], dim=c(3,4))
legend("topright",legend=levels(factor(targets$Sample_Group)),text.col=pal)
```

### Filtering
Poor performing probes are generally filtered out prior to differential methylation analysis. As the signal from these probes is unreliable, by removing them we perform fewer statistical tests and thus incur a reduced multiple testing adjustment. We filter out probes that have failed in one or more samples based on detection p-value. The detection  p-values are automatically calculated using the signal from the negative control probes on the array; probes that are determined to not have a signal significantly different from the background level have high detection p-values.   
```{r, cache=TRUE}
## remove any probes that have failed in one or more samples
detP = detP[match(featureNames(mSetSq),rownames(detP)),] # ensure probes are in the same order in the mSetSq and detP objects
keep = rowSums(detP < 0.01) == ncol(mSetSq) 
table(keep)

mSetSqFlt = mSetSq[keep,]
mSetSqFlt
```

Depending on the nature of your samples and your biological question you may also choose to filter out the probes from the X and Y chromosomes or probes that are known to have common SNPs at the CpG site. This can be done using the code below, however, we will not be filtering out these probes as part of the analysis in this workshop.
```{r, eval=FALSE}
## if your data infludes males and females, remove the sex chromosomes
keep = !(featureNames(mSetSqFlt) %in% ann450k$Name[ann450k$chr %in% c("chrX","chrY")])
table(keep)

mSetSqFlt = mSetSqFlt[keep,]

## remove probes with SNPs at CpG or SBE site
mSetSqFlt = dropLociWithSnps(mSetSqFlt)
```

We will also filter out probes that have shown to be cross-reactive i.e. map to multiple places in the genome. This list was oroginally published by [Chen et al.](http://www.tandfonline.com/doi/abs/10.4161/epi.23470) in 2013 and can be downloaded directly from the author's [website](http://www.sickkids.ca/MS-Office-Files/Research/Weksberg%20Lab/48639-non-specific-probes-Illumina450k.xlsx). The Excel spreadsheet containing the cross-reactive probes is also available as part of the data download for this workshop.
```{r, cache=TRUE}
## exclude cross reactive probes 
xReactiveProbes = read.csv(file=paste(dataDirectory,"48639-non-specific-probes-Illumina450k.csv",sep="/"), stringsAsFactors=FALSE)
keep = !(featureNames(mSetSqFlt) %in% xReactiveProbes$TargetID)
table(keep)

mSetSqFlt = mSetSqFlt[keep,] 
mSetSqFlt
```

After the data has been normalised and filtered, we can calculate M and $\beta$ values. $$M=log2(methylated/unmethylated)$$ $$\beta=methylated/(methylated+unmethylated)$$ M-values have nicer statistical properties and are thus better for use in statistical analysis of methylation data whilst $\beta$ values are easy to interpret and are thus better for displaying data. A detailed comparison of M and $\beta$ values was published by [Du et al.](http://www.biomedcentral.com/1471-2105/11/587) in 2010.  
```{r, cache=TRUE}
## calculate M-values for statistical analysis
mVals = getM(mSetSqFlt)
head(mVals)

bVals = getBeta(mSetSqFlt)
head(bVals)

par(mfrow=c(1,2))
densityPlot(bVals,sampGroups = targets$Sample_Group,main="Beta values")
densityPlot(mVals,sampGroups = targets$Sample_Group,main="M values")
```


### Differential methylation analysis (probes)
The primary biological goal for this dataset is to discover differentially methylated probes between the different cell types. However, there is another factor that we need to take into account when we perform the modelling. In the `targets` file, there is a column called `Sample_Source`, and this refers to the individuals from whom the samples came. In this dataset, any one particular individual contributes more than one cell type. For example, individual M28 contributes naive, rTreg and actnaive samples. From the MDS plots we know that the samples cluster according to individual first, and then by cell type, implying that individual is the greatest source of variation in methylation. In order to get the cell type differences, we need to take into account individual variation. Hence when we specify our design matrix, we need to include two factors: individual and cell type. This style of analysis is called a paired analysis; differences between cell types are calculated *within* each individual, and then these differences are summed *across* individuals to determine whether overall there is a significant difference in the mean methylation level for each CpG site.
```{r, cache=TRUE}
cellType = factor(targets$Sample_Group) ## this is the factor of interest
individual = factor(targets$Sample_Source) ## this is the individual effect that we need to account for

## use the above to create a design matrix
design = model.matrix(~0+cellType+individual, data=targets)
colnames(design) = c(levels(cellType),levels(individual)[-1])
  
fit = lmFit(mVals, design) ## fit the linear model
## create a contrast matrix for specific comparisons
contMatrix = makeContrasts(naive-rTreg,
                           naive-act_naive,
                           rTreg-act_rTreg,
                           act_naive-act_rTreg,
                           levels=design)
contMatrix

## fit the contrasts
fit2 = contrasts.fit(fit, contMatrix)
fit2 = eBayes(fit2)

## look at the numbers of DM CpGs at FDR < 0.05
summary(decideTests(fit2))

## get the table of results for the first contrast (naive - rTreg)
ann450kSub = ann450k[match(rownames(mVals),ann450k$Name),c(1:4,12:19,24:ncol(ann450k))]
DMPs = topTable(fit2, num=Inf, coef=1, genelist = ann450kSub)
head(DMPs)

```
The results of the analysis for the first comparison, naive vs rTreg, are stored in the `DMPs` object. We can write out this file as a comma separated file, which can be opened in excel.
```{r, eval=FALSE}
write.table(DMPs, file="DMPs.csv", sep=",", row.names=FALSE)

```
It is always useful to plot the top differentially methylated CpG sites - it serves as a sanity check to make sure the results make sense, and if they don't it usually means there is an error in the code somewhere, or perhaps the design matrix was not set up correctly. It is easier to interpret methylation on the beta value scale, so although we always perform analysis on the Mvalue scale, we usually visualise data on the beta value scale.
```{r, cache=TRUE}
par(mfrow=c(2,2))
sapply(rownames(DMPs)[1:4], function(cpg){
  plotCpg(bVals, cpg=cpg, pheno = targets$Sample_Group)
})

```

# Additional analyses
## Gene ontology testing

Once you have performed a differential methylation analysis, there may be a very long list of significant CpG sites to interpret. One question a researcher may have is which gene pathways may be targeted by the significant CpGs. Sometimes it is easy to link the top differentially methylated CpGs to genes that make sense in terms of the cell types/samples being studied, but there can be many more thousands of CpGs significantly differentially methylated. In order to get an idea of the biological processes that the significant CpGs may be involved in, we can perform gene ontology testing with the `gometh` function in the *missMethyl* package.

Let us consider the first comparison, naive vs rTreg, with the results of the analysis in the `DMPs` table. The `gometh` function takes as input a character vector of the names (e.g. cg20832020) of the significant CpG sites, and optionally, a character vector of all CpGs tested. This is recommended if you have performed quite a lot of filtering of the CpGs before analysis. In the `DMPs` table, the `Name` column corresponds to the CpG name. We will select all CpG sites that have adjusted p-value of less than 0.05.

```{r, cache=TRUE}
## Get the significant CpG sites at less than 5% FDR
sigCpGs = DMPs$Name[DMPs$adj.P.Val<0.05]
## First 10 significant CpGs
sigCpGs[1:10]
## Total number of significant CpGs at 5% FDR
length(sigCpGs)
## Get all the CpG sites used in the analysis to form the background
all = DMPs$Name
## Total number of CpG sites tested
length(all)
```

What makes `gometh` work particularly well with methylation data is that it can take into account the varying numbers of CpGs associated with genes. For the 450K array, the numbers of CpGs mapping to genes can vary from as few as 1 to as many as 1200. The genes that have more CpGs associated with them will have a higher probability of being identified as differentially methylated compared to genes with fewer CpGs. We can look at the bias in the data by specifying `plot=TRUE` in the call to `gometh`.

```{r, cache=TRUE}
par(mfrow=c(1,1))
gst <- gometh(sig.cpg=sigCpGs, all.cpg=all, plot.bias=TRUE)
```

The `gst` object is a dataframe with each row corresponding to the GO category being tested. The top 20 gene ontology categories can be displayed using the `topGO` function:

```{r, cache=TRUE}
## Top 20 GO categories
topGO(gst)
```

Here we can see lots of GO categories corresponding to immune system and T cell processes, which makes sense as the cell types being studied form part of the immune system.
The total number of significant GO categories, adjusted for multiple testing, is:

```{r, cache=TRUE}
## Total number of significant GO categories at 5% FDR
sum(gst$FDR<0.05)
```

## Differential variability
Rather than testing for differences in mean methylation, we may be interested in testing for differences between group variances. For example, it has been hypothesised that highly variable CpGs in cancer are important for tumour progression. Hence we may be interested in CpG sites that are consistently methylated in one group, but variably methylated in another group. In general we recommend at least 10 samples in each group for testing differential variability in order to get a good estimate of the variance. It is more tricky to accurately estimate variances as opposed to means, so you generally need a larger sample size for testing differences in group variances as opposing to testing differences in group means. For the purpose of this workshop, we won't worry about this sample size issue.

The way we specify which groups we are interested in testing is a little bit different to the way you specify a model in `limma` for differential methylation, particularly if you fit an intercept model. For our data, we have set up a design matrix without an intercept term, and we will specify which groups we want to test for differential variability in a contrasts statement. We can use the same design matrix that we set up for testing differential methylation.
```{r, cache=TRUE}
## Fit the model for differential variability, specifying that the four cell types are the groups of interest
fitvar <- varFit(mVals, design = design, coef = c(1,2,3,4))
## Specify the groups we are interested in testing for differential variability
contr <- makeContrasts(naive-rTreg,
                       naive-act_naive,
                       rTreg-act_rTreg,
                       act_naive-act_rTreg,
                       levels=design)
fitvar <- contrasts.varFit(fitvar,contrasts=contr)
## Summary of differential variability
summary(decideTests(fitvar))
topDV <- topVar(fitvar, coef=1)
## Top 10 differentially variable CpGs between naive and rTreg
topDV
```
Just as we had a look at the beta values for the significant differentially methylated CpGs, it is useful to plot differentially variable CpGs too.

```{r, cache=TRUE}
par(mfrow=c(2,2))
sapply(rownames(topDV)[1:4], function(cpg){
  plotCpg(bVals, cpg=cpg, pheno = targets$Sample_Group)
})
```
An example of testing for differential variability when you have a design matrix with an intercept term is detailed in the *missMethyl* [vignette]("http://www.bioconductor.org/packages/release/bioc/vignettes/missMethyl/inst/doc/missMethyl.pdf").

## Differential methylation analysis (regions)
Although performing a *probe-wise* analysis is useful and informative, sometimes we are interested in knowing whether several proximal CpGs are concordantly differentially methylated i.e. we want to identify differentially methylated **regions**. There are several Bioconductor packages that can identify differentially methylated regions from 450k data. Some of the most popular are [DMRcate](https://www.bioconductor.org/packages/release/bioc/html/DMRcate.html), [minfi](http://bioconductor.org/packages/release/bioc/html/minfi.html) plus [bumphunter](https://www.bioconductor.org/packages/3.3/bioc/html/bumphunter.html) and [charm](http://www.bioconductor.org/packages/release/bioc/html/charm.html). They are each based on different statistical methods. In this workshop, we are using *DMRcate* because it is faster than the other methods and is based on *limma*, so we can directly use the `design` and `contMatrix` we previously defined.

Firstly, our matrix of M-values is annotated with relevant information about the probes such as their genomic position, gene annotation, etc. By default, this is done using the `ilmn12.hg19` annotation, but this can be substituted for any argument compatible with the interface provided by the *minfi* package. The backbone of the *limma* pipeline is then used for differential array analysis to calculate t-statistics and, optionally, filter probes by their FDR-corrected p-value.
```{r, cache=TRUE}
myAnnotation = cpg.annotate(mVals, analysis.type="differential", design=design, contrasts = TRUE,
                             cont.matrix = contMatrix, coef="naive - rTreg")

str(myAnnotation)
```

We can then use the `dmrcate` function to look for differentially methylated regions. The main output table `DMRs$results` contains all of the regions found, along with their genomic annotations and p-values.
```{r, cache=TRUE}
DMRs = dmrcate(myAnnotation, lambda=1000, C=2)
head(DMRs$results)
```

The regions can be quickly examined using the `DMR.plot` function provided in the *DMRcate* package. 
```{r, cache=TRUE}
par(mfrow=c(1,1))
DMR.plot(dmrcoutput=DMRs, dmr=1, betas=bVals, phen.col=pal[factor(targets$Sample_Group)],
         pch=16, toscale=TRUE, plotmedians=TRUE)
legend("bottomleft",legend=levels(factor(targets$Sample_Group)),col=pal,lty=1)

```

# Visualisations
The *Gviz* package offers some powerful functionality for plotting genomic data in its genomic context. Their extensive [vignette](https://bioconductor.org/packages/release/bioc/vignettes/Gviz/inst/doc/Gviz.pdf) covers the various types of plots that can be produced using their framework. We will re-plot the top differentially methylated region from the previous analysis to demonstrate the richness of the visualisations that can be created.

Firstly, we will set up the genomic region we would like to plot by extracting the genomic coordinates of the top differentially methylated region.

```{r, cache=TRUE}
gen = "hg19"
coords = strsplit2(DMRs$results$hg19coord[1],":")
chrom = coords[1]
minbase = as.numeric(strsplit2(coords[2],"-")[1]) - 2000
maxbase = as.numeric(strsplit2(coords[2],"-")[2]) + 2000
```

Next we will load up some genomic annotations of interest such as the locations of CpG islands and DNAseI hypersensitive sites.
```{r, cache=TRUE}
## CpG islands
islandHMM = read.csv(paste(dataDirectory,"model-based-cpg-islands-hg19.txt",sep="/"), 
                     sep="\t", stringsAsFactors =FALSE, header=TRUE)
head(islandHMM)

islandData = GRanges(seqnames = Rle(islandHMM$chr), 
                     ranges = IRanges(islandHMM$start, end = islandHMM$end),
                    strand = Rle(strand(rep("*",nrow(islandHMM)))))
islandData = islandData[seqnames(islandData) == chrom & 
                         (start(islandData) >= minbase & end(islandData) <= maxbase)] # this is to reduce the amount of data in memory
islandData

## DNAseI hypersensitive sites
dnase = read.csv(paste(dataDirectory,"wgEncodeRegDnaseClusteredV3.bed",sep="/"),
                 sep="\t",stringsAsFactors=FALSE,header=FALSE)
head(dnase)

dnaseData = GRanges(seqnames = dnase[,1],
                    ranges = IRanges(dnase[,2], end = dnase[,3]),
                    strand = Rle(rep("*",nrow(dnase))),
                    data = dnase[,5])
dnaseData = dnaseData[seqnames(dnaseData) == chrom & 
                          (start(dnaseData) >= minbase & end(dnaseData) <= maxbase)] # this is to reduce the amount of data in memory
dnaseData
```

Now, set up the ideogram, genome and RefSeq tracks that will provide context for our methylation data. 
```{r, cache=TRUE}
iTrack = IdeogramTrack(genome = gen, chromosome = chrom, name="")
gTrack = GenomeAxisTrack(col="black", cex=1, name="", fontcolor="black")
rTrack = UcscTrack(genome = gen, chromosome = chrom, track = "refGene", from = minbase, 
                            to = maxbase, trackType = "GeneRegionTrack", rstarts = "exonStarts", 
                            rends = "exonEnds", gene = "name", symbol = "name2", transcript = "name", 
                            strand = "strand", fill = "darkblue",stacking = "squish", name = "RefSeq", 
                            showId=TRUE, geneSymbol=TRUE)
```

Ensure that the methylation data is ordered by chromosome and base position.
```{r, cache=TRUE}
ann450kOrd = ann450kSub[order(ann450kSub$chr,ann450kSub$pos),]
head(ann450kOrd)

bValsOrd = bVals[match(ann450kOrd$Name,rownames(bVals)),]
head(bValsOrd)
```

Create the data tracks using the appropriate track type for each data type.
```{r, cache=TRUE}

methTrack = DataTrack(start=ann450kOrd$pos, width=1, 
                      data=t(bValsOrd), 
                     groups=targets$Sample_Group,genome = gen,chromosome = chrom, ylim=c(-0.05,1.05),
                     col=pal,type=c("a","p"), name="DNA Meth.\n(beta value)", background.panel="white",legend=TRUE,
                     cex.title=0.4,cex.axis=0.5,cex.legend=0.8)
islandTrack = AnnotationTrack(range=islandData, genome = gen, name = "CpG Is.", fill="grey40",
                              chromosome = chrom)
dnaseTrack = DataTrack(range=dnaseData, genome = gen, name = "DNAseI", type="gradient",
                        chromosome = chrom)
```

Set up the track list and indicate the relative sizes of the different tracks. Finally, make the plot using the `plotTracks` function.
```{r, cache=TRUE}
tracks = list(iTrack,gTrack,methTrack,islandTrack,dnaseTrack,rTrack)
sizes = c(3,3,5,3,3,5)

plotTracks(tracks, from = minbase, to = maxbase, showTitle = TRUE, add53 = TRUE, add35 = TRUE, 
             grid=TRUE, lty.grid=3, sizes=sizes, length(tracks))


```

# Software versions
```{r}
sessionInfo()
```
